\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{url}
\usepackage{balance}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc}

\definecolor{codegreen}{rgb}{0.1,0.5,0.1}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codeblue}{rgb}{0.0,0.2,0.6}
\definecolor{backgray}{rgb}{0.96,0.96,0.96}

\lstdefinestyle{terminal}{
  backgroundcolor=\color{backgray},
  basicstyle=\ttfamily\scriptsize,
  breaklines=true,
  frame=single,
  rulecolor=\color{codegray},
  keywordstyle=\color{codeblue},
  commentstyle=\color{codegreen},
  numbers=none,
  xleftmargin=4pt,xrightmargin=4pt,
  aboveskip=4pt,belowskip=4pt
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Self-Hosted LLM Orchestration for Autonomous Penetration Testing\\
with Three-Layer Scope Enforcement and\\
Tamper-Evident Cryptographic Audit Chains}

\author{%
  \IEEEauthorblockN{Md Anisur Rahman Chowdhury\textsuperscript{},
                    Ronny Bazan-Antequera\textsuperscript{}}
  \IEEEauthorblockA{%
    \textsuperscript{}Dept.\ of Computer and Information Science,
    Gannon University, USA\\
    \texttt{engr.aanis@gmail.com,\; bazanant001@gannon.edu}}
}

\maketitle

%─────────────────────────────────────────────────────────────
\begin{abstract}
Penetration testing is a labour-intensive discipline that demands expert judgement
at every step of the assessment lifecycle. Existing automation tools cover
individual scan tasks well, but none offer a coherent, scope-safe reasoning layer
capable of chaining observations into a multi-phase attack narrative.
This paper presents \textbf{PentAI Pro}, a platform that orchestrates
self-hosted large language models (LLMs) across physically isolated
virtual machines to automate the full penetration-testing pipeline:
Reconnaissance, Enumeration, Vulnerability Mapping, and Exploitation Planning.
A three-layer scope-enforcement architecture independently validates every
planned and executed action at the LLM planning stage, the backend API,
and the remote tool gateway, yielding 100\,\% out-of-scope rejection in
all tested scenarios.
A SHA-256 hash-chained event log provides cryptographically verifiable
evidence of all system actions.
All inference is handled by a locally deployed 14-billion-parameter
open-weight model (\textit{qwen2.5:14b-instruct}), eliminating data-egress
risks inherent in cloud-hosted inference.
Evaluation on a three-VM Proxmox laboratory reports a mean reconnaissance
planning latency of 18.4\,s, full-pipeline completion under 4\,min,
and a 108-test integration suite that passes in 43.5\,s.
\end{abstract}

\begin{IEEEkeywords}
penetration testing automation, large language models,
retrieval-augmented generation, scope enforcement, mTLS,
tamper-evident audit, self-hosted AI, security orchestration
\end{IEEEkeywords}

%─────────────────────────────────────────────────────────────
\section{Introduction}

Security assessments are commissioned to find weaknesses before attackers
do. A human practitioner working a typical external assessment spends
time on reconnaissance, port scanning, service enumeration, and
vulnerability correlation — largely mechanical work punctuated by
occasional insight. As the scale of modern infrastructure grows,
the mechanical fraction of that work has become a bottleneck.

Scan-based tools (Nessus, OpenVAS, Nuclei) handle the mechanical steps
efficiently but produce flat lists of findings rather than coherent
attack narratives. They do not reason about how a misconfigured TLS
setting and an outdated web framework on the same host combine into an
exploitable path. Filling that gap has, until recently, required a human.

Large language models introduce a new option. Models trained on security
documentation, CVE descriptions, and published offensive research
carry enough domain knowledge to propose multi-step attack plans,
interpret raw scanner output, and draft professional findings.
The central challenge is not whether such a model can reason about
offensive security — recent work has confirmed it can~\cite{happe2023,deng2023}.
The challenge is how to deploy that reasoning safely: respecting engagement
scope, preserving an immutable record of every action, and preventing
adversarially crafted target responses from manipulating the agent.

This paper addresses those challenges through three concrete contributions:

\begin{enumerate}
\item A \textbf{multi-phase LLM agent pipeline} integrating a
  retrieval-augmented knowledge base, phase-aware planning nodes,
  and a typed tool registry — all running on-premises.
\item A \textbf{three-layer scope-enforcement architecture} that
  independently validates every action at the LLM planning,
  backend API, and weapon-gateway levels.
\item A \textbf{SHA-256 hash-chained audit log} that links every
  security event into a forensically verifiable chain.
\end{enumerate}

We describe the design, implementation, and experimental evaluation of
PentAI Pro, deployed on a three-VM Proxmox laboratory with real pentest
tools (nmap, Nuclei, Gobuster, httpx, sslscan).
The remainder of this paper is organised as follows.
Section~II surveys related work.
Section~III presents the system architecture.
Section~IV describes the implementation.
Section~V reports experimental results.
Section~VI discusses implications and limitations, and
Section~VII concludes.

%─────────────────────────────────────────────────────────────
\section{Background and Related Work}

\subsection{Scan-Based Automation}

Nessus~\cite{nessus} and OpenVAS follow a plug-in model: each plug-in
tests one condition, and the aggregated results form a vulnerability
report. Nuclei~\cite{nuclei} generalises this with community-maintained
YAML templates. These tools excel at breadth but have no mechanism
for reasoning about the interdependence of findings or planning follow-on
steps based on prior observations.

\subsection{LLMs Applied to Offensive Security}

Happe and Cito~\cite{happe2023} ran GPT-4 in a penetration-testing loop,
showing measurable progress on capture-the-flag challenges.
PentestGPT~\cite{deng2023} introduced a multi-turn dialogue between
the operator and an LLM planner; operator interaction was needed at
every decision point.
PentAI Pro removes that per-step dependency: the operator defines scope
and goal once; the agent executes autonomously, surfacing only
high-risk approval requests.
ReAct~\cite{yao2022} formalises reasoning-action-observation loops for
general LLM agents. PentAI Pro specialises this pattern with
typed tool schemas, phase-aware supervisors, and prompt-injection defences
absent from general-purpose frameworks.

\subsection{Retrieval-Augmented Generation}

Lewis et al.~\cite{lewis2020} showed that conditioning generation on
retrieved passages significantly improves factual accuracy for knowledge-intensive
tasks. Penetration testing is knowledge-intensive: the quality of a
reconnaissance plan depends on knowing which nmap flags reveal OS fingerprints,
which HTTP headers indicate vulnerable frameworks, and which CVEs affect
detected service versions. PentAI Pro uses a PostgreSQL-backed vector store
with nomic-embed-text embeddings to retrieve and inject relevant playbook
fragments into every planning prompt.

\subsection{Security Orchestration Platforms}

SOAR platforms (e.g., Splunk SOAR, Palo Alto XSOAR) automate defensive
workflows: alert triage, indicator enrichment, and incident response runbooks.
Offensive automation at a comparable level of orchestration — with scope
enforcement, inter-node mTLS, and cryptographic audit trails — has not been
reported in published literature targeting fully autonomous operation.
The closest adjacent work in offensive automation focuses on single-tool
wrappers or one-shot exploit generation~\cite{andriesse2018},
rather than on the multi-phase, scope-aware agent coordination that
characterises PentAI Pro.
The OWASP Testing Guide~\cite{owasp} defines the methodological baseline
that informed our phase decomposition (Reconnaissance, Enumeration,
Vulnerability Identification, Exploitation): PentAI Pro's four-phase
pipeline maps directly onto Sections 4--7 of that guide.

%─────────────────────────────────────────────────────────────
\section{System Architecture}

\subsection{Infrastructure}

PentAI Pro runs across three isolated virtual machines on a Proxmox
hypervisor connected by a private VLAN (172.20.32.0/18), as shown in
Fig.~\ref{fig:arch}. Physical isolation ensures that a compromise of the
Weapon Node cannot directly reach the Command Node's database or LLM
inference engine.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=1.1cm and 1.4cm,
  box/.style={draw,rounded corners=2pt,minimum width=2.2cm,
              minimum height=0.55cm,font=\scriptsize,align=center},
  vm/.style={draw,thick,rounded corners=3pt,inner sep=6pt},
  arr/.style={-Stealth,thick},
  mtls/.style={-Stealth,thick,dashed,color=blue!60}
]
% Command Node
\node[vm,fill=blue!6,minimum width=3.6cm,minimum height=2.8cm] (cmd) {};
\node[above=0pt of cmd.north,font=\scriptsize\bfseries,yshift=-2pt]
  {Command Node (172.20.32.74)};
\node[box,fill=blue!15] (fe) at ([xshift=0pt,yshift=0.7cm]cmd.center)
  {Next.js\\Frontend};
\node[box,fill=blue!15,below=0.25cm of fe] (be) {FastAPI\\Backend};
\node[box,fill=blue!15,below=0.25cm of be] (db) {PostgreSQL + Ollama};

% Weapon Node
\node[vm,fill=orange!6,minimum width=2.6cm,minimum height=1.2cm,
      right=1.6cm of cmd] (wpn) {};
\node[above=0pt of wpn.north,font=\scriptsize\bfseries,yshift=-2pt]
  {Weapon Node (.68)};
\node[box,fill=orange!15] (gw) at (wpn.center)
  {Tool Gateway\\(mTLS :5000)};

% Target Node
\node[vm,fill=red!6,minimum width=2.2cm,minimum height=0.9cm,
      below=1.0cm of wpn] (tgt) {};
\node[font=\scriptsize\bfseries] at (tgt.north)
  [yshift=-8pt]{Target (.59)};

% Arrows
\draw[arr] (fe) -- (be);
\draw[arr] (be) -- (db);
\draw[mtls] (be.east) -- node[above,font=\tiny,color=blue!70]{JWT+mTLS}
  (gw.west);
\draw[arr,color=red!60] (gw.south) -- node[right,font=\tiny]{scan} (tgt.north);
\end{tikzpicture}
\caption{Three-VM deployment on Proxmox. Traffic between the Backend
  and Tool Gateway is secured by mutual TLS and short-lived scoped JWTs.}
\label{fig:arch}
\end{figure}

\begin{table}[t]
\caption{Virtual Machine Roles and Specifications}
\centering\small
\begin{tabular}{lllp{1.8cm}}
\toprule
Node & IP & OS & Hardware\\
\midrule
Command & 172.20.32.74 & Ubuntu 22.04 & 6-core Xeon, 25\,GB RAM, 400\,GB SSD\\
Weapon  & 172.20.32.68 & Kali Linux   & 4-core, 8\,GB RAM\\
Target  & 172.20.32.59 & Configurable & 2-core, 4\,GB RAM\\
\bottomrule
\end{tabular}
\label{tab:vms}
\end{table}

\subsection{Three-Layer Scope Enforcement}

Scope enforcement is the safety-critical property of the system.
Any failure — a planned target outside the engagement's CIDR, or a tool
execution against an unscoped host — constitutes an unauthorised
computer-access offence in most jurisdictions.
PentAI Pro enforces scope at three independent points so that no single
failure can cause an out-of-scope action:

\noindent\textbf{Layer~1 — LLM Planning.}
The planning prompt provides the engagement's \texttt{scope\_cidrs} as
an explicit constraint. Every target address in the model's structured
JSON output is validated against these CIDRs before any downstream step
proceeds. Proposals naming addresses outside the scope are discarded
silently.

\noindent\textbf{Layer~2 — Backend API.}
The FastAPI middleware re-validates CIDR membership for every tool
invocation, independently of the planner's prior check.
High-risk operations (Nuclei targeted scans, Gobuster directory
enumeration, OS detection) additionally require a persisted operator
approval before a JWT is issued to the Weapon Node.

\noindent\textbf{Layer~3 — Weapon Gateway.}
The Flask gateway on the Weapon Node verifies the JWT audience,
signature, and the scope field embedded in the token on every
incoming request — regardless of what either previous layer accepted.

\subsection{Tamper-Evident Audit Chain}

Every security event is appended to a SHA-256 hash-chained log:
\begin{equation}
h_n = \mathrm{SHA256}\!\bigl(\,\mathrm{JSON}(e_n)\,\|\,h_{n-1}\bigr),
\quad h_0 = \texttt{0}^{64}
\label{eq:chain}
\end{equation}
where $e_n$ contains the event type, structured payload, UTC timestamp,
and actor identifier.
Re-computing the chain from $h_0$ detects any modification, insertion,
or deletion; the first mismatch identifies the tampered position.

%─────────────────────────────────────────────────────────────
\section{Implementation}

\subsection{Agent Pipeline}

The supervisor orchestrates six sequential nodes, shown in
Algorithm~\ref{alg:pipeline}. Each node calls the LLM, validates the
response against a typed schema, and passes the result to the next node.
Phase selection depends on the classified intent so that a
\texttt{recon\_only} engagement skips enumeration and exploitation nodes.

\begin{algorithm}[t]
\caption{PentAI Pro Agent Pipeline}
\label{alg:pipeline}
\small
\begin{algorithmic}[1]
\REQUIRE engagement $E$ with $\texttt{scope\_cidrs}$, operator goal $g$
\ENSURE structured findings $F$, step results $R$
\STATE $\textit{intent} \leftarrow \mathrm{Classify}(g)$ \COMMENT{3B model}
\STATE $\textit{plan}_r \leftarrow \mathrm{PlanRecon}(g,\,E,\,\mathrm{KB})$
  \COMMENT{14B model + RAG}
\STATE $R \leftarrow R \cup \mathrm{Execute}(\textit{plan}_r,\,E)$
  \COMMENT{Weapon Node via mTLS}
\IF{$\textit{intent} \in \{\texttt{recon\_enum},\,\texttt{full\_pentest}\}$}
  \STATE $\textit{plan}_e \leftarrow \mathrm{PlanEnum}(R,\,E,\,\mathrm{KB})$
  \STATE $R \leftarrow R \cup \mathrm{Execute}(\textit{plan}_e,\,E)$
\ENDIF
\IF{$\textit{intent} \in \{\texttt{full\_pentest},\,\texttt{targeted}\}$}
  \STATE $F \leftarrow \mathrm{MapVulnerabilities}(R,\,\mathrm{KB})$
  \STATE $F \leftarrow F \cup \mathrm{PlanExploitation}(F,\,E)$
\ENDIF
\RETURN $F,\,R$
\end{algorithmic}
\end{algorithm}

\subsection{LLM Model Routing}

Two open-weight models serve different roles based on task
computational requirements:

\begin{itemize}
\item \textbf{qwen2.5:14b-instruct-q4\_K\_M} (14B params, 4-bit
  quantised, $\approx$9\,GB RAM). Planning, vulnerability mapping,
  finding generation: tasks requiring extended reasoning over complex,
  heterogeneous context. Achieves $\approx$12\,tok/s on the 6-core Xeon.
\item \textbf{llama3.2:3b-instruct-q4\_K\_M} (3B params, $\approx$2\,GB
  RAM). Intent classification and raw tool output parsing: structured,
  short-context tasks where latency matters more than depth.
  Achieves $\approx$45\,tok/s on the same hardware.
\item \textbf{nomic-embed-text} (137M params, 274\,MB). Generates
  768-dimensional embeddings for the RAG knowledge base.
\end{itemize}

\subsection{Retrieval-Augmented Knowledge Base}

Security playbooks and methodology notes are ingested as markdown
documents, chunked at heading boundaries (200–1800 characters), and
embedded with nomic-embed-text. At planning time, the top-$k=4$ chunks
above a cosine-similarity threshold of 0.15 are prepended to the
planning prompt. Source citations are tracked in each plan step.

\subsection{Prompt Injection Defence}

Tool output from the Weapon Node is data from a potentially adversarial
source. A compromised target could serve crafted nmap banners or HTTP
headers designed to alter the agent's subsequent reasoning.
PentAI Pro wraps all tool output in trust-annotated XML before the
LLM receives it:

\begin{lstlisting}[style=terminal,caption={Tool output sandboxing}]
<tool_output trust="untrusted">
Starting Nmap 7.95 ...
PORT   STATE SERVICE VERSION
80/tcp open  http    Apache 2.4.49
443/tcp open  ssl/http nginx 1.18.0
</tool_output>
\end{lstlisting}

The system prompt instructs the model to treat
\texttt{trust="untrusted"} content as raw bytes to be parsed, never
as instructions. This structural separation is not a complete defence
against all injection forms, but it substantially raises the bar for
indirect injection via tool output.

\subsection{Real-Time Execution Streaming}

Tool executions on the Weapon Node produce NDJSON event streams.
An in-process \texttt{ExecutionBus} fans these events to all active
WebSocket subscribers using per-subscriber asyncio queues.
A thread-safe bridge (\texttt{loop.call\_soon\_threadsafe}) connects the
synchronous HTTPS streaming reader to the async event loop.
Slow subscribers are dropped (queue full) rather than allowed to
block the producer. Measured end-to-end latency from Weapon Node to
browser was consistently below 50\,ms.

A representative live stream from a service-scan execution:

\begin{lstlisting}[style=terminal,caption={Real-time NDJSON stream (truncated)}]
{"type":"timeout_seconds","value":120}
{"type":"stdout","line":"Starting Nmap 7.95..."}
{"type":"stdout","line":"Nmap scan report for 172.20.32.59"}
{"type":"stdout","line":"PORT    STATE  SERVICE  VERSION"}
{"type":"stdout","line":"22/tcp  open   ssh      OpenSSH 8.9p1"}
{"type":"stdout","line":"80/tcp  open   http     Apache httpd 2.4.52"}
{"type":"stdout","line":"443/tcp open   ssl/http  nginx 1.18.0"}
{"type":"completed","status":"completed","exit_code":0}
\end{lstlisting}

%─────────────────────────────────────────────────────────────
\section{Experimental Evaluation}

\subsection{Setup}

All experiments ran on the Command Node (Intel Xeon Gold 6230,
6\,vCPUs at 2.10\,GHz, 25\,GB RAM, no discrete GPU).
The Weapon Node (Kali Linux 2024.1) ran nmap~7.95, Nuclei~3.3,
Gobuster~3.6, and httpx~1.6.
Each benchmark trial used one in-scope target host at 172.20.32.59.
Thirty independent trials were collected per metric;
means and standard deviations are reported throughout.

\subsection{Phase Latency}

Fig.~\ref{fig:latency} shows the mean execution time of each pipeline
phase. Planning nodes (14B model) dominate latency; classification
(3B model) is sub-2\,s. Knowledge-base retrieval adds a flat 0.8\,s
regardless of query complexity.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  xbar,
  width=\columnwidth,height=5.2cm,
  bar width=6pt,
  xlabel={Time (seconds)},
  symbolic y coords={
    KB Retrieval,
    Classify,
    Write Finding,
    Plan Enum,
    Plan Recon,
    Map Vulns,
    Plan Exploit},
  ytick=data,
  yticklabel style={font=\scriptsize},
  xticklabel style={font=\scriptsize},
  xlabel style={font=\scriptsize},
  nodes near coords,
  nodes near coords style={font=\tiny,xshift=2pt},
  every node near coord/.append style={/pgf/number format/.cd,fixed,precision=1},
  xmin=0,xmax=38,
  enlarge y limits=0.12,
  grid=major,grid style={dotted,gray!40},
  legend pos=south east,
  legend style={font=\scriptsize},
]
\addplot+[fill=blue!40,draw=blue!60] coordinates {
  (0.8,KB Retrieval)
  (1.2,Classify)
  (11.2,Write Finding)
  (15.1,Plan Enum)
  (18.4,Plan Recon)
  (22.3,Map Vulns)
  (28.1,Plan Exploit)
};
\end{axis}
\end{tikzpicture}
\caption{Mean phase latency (30 trials, error bars = 1 SD).
  KB retrieval is nearly constant; exploitation planning shows the
  highest variance due to variable finding depth.}
\label{fig:latency}
\end{figure}

\subsection{Tool Execution Times}

Fig.~\ref{fig:tools} compares mean execution times across the
eight tool operations in the registry. httpx and dnsx complete in
under 8\,s; Nuclei targeted scans take the longest at 78.4\,s
due to template enumeration.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  width=\columnwidth,height=4.6cm,
  bar width=7pt,
  ylabel={Time (seconds)},
  symbolic x coords={
    dnsx, httpx, whatweb, sslscan,
    nmap-top, nmap-svc, nmap-os, nuclei},
  xtick=data,
  xticklabel style={font=\tiny,rotate=30,anchor=east},
  yticklabel style={font=\scriptsize},
  ylabel style={font=\scriptsize},
  ymin=0,ymax=105,
  nodes near coords,
  nodes near coords style={font=\tiny,rotate=90,anchor=west},
  every node near coord/.append style={/pgf/number format/.cd,fixed,precision=0},
  grid=major,grid style={dotted,gray!40},
  enlarge x limits=0.08,
]
\addplot+[fill=orange!50,draw=orange!70] coordinates {
  (dnsx,   4.9)
  (httpx,  5.8)
  (whatweb,9.1)
  (sslscan,13.8)
  (nmap-top,31.4)
  (nmap-svc,42.1)
  (nmap-os, 58.4)
  (nuclei,  78.4)
};
\end{axis}
\end{tikzpicture}
\caption{Mean tool execution times (30 trials per tool, error bars = 1 SD).
  Nuclei's high variance reflects variable template match counts.}
\label{fig:tools}
\end{figure}

\subsection{Scope Enforcement Correctness}

We designed three adversarial scenarios to verify each enforcement layer:
(S1) crafted LLM output proposing an out-of-scope target,
(S2) a direct backend API call bypassing the LLM, and
(S3) a forged JWT sent directly to the Weapon Node gateway.
Table~\ref{tab:scope} shows results across 50 trials per scenario.

\begin{table}[t]
\caption{Scope Enforcement Verification (50 trials each)}
\centering\small
\begin{tabular}{llrr}
\toprule
Scenario & Blocked at & Blocked & Escaped\\
\midrule
S1: Crafted LLM output  & Layer 1 (planner) & 50/50 & 0\\
S2: Direct API bypass   & Layer 2 (backend) & 50/50 & 0\\
S3: Forged JWT to gateway & Layer 3 (gateway) & 50/50 & 0\\
\midrule
\textbf{Overall} & & \textbf{150/150} & \textbf{0}\\
\bottomrule
\end{tabular}
\label{tab:scope}
\end{table}

\subsection{Audit Chain Integrity}

Table~\ref{tab:audit} measures chain-validation time and tamper-detection
reliability across three chain lengths. A single event was modified at a
random position in each trial; the validator identified the exact position
in all cases.

\begin{table}[t]
\caption{Audit Chain Validation Performance (10 trials each)}
\centering\small
\begin{tabular}{lrrr}
\toprule
Chain Length & Validation (ms) & Tamper Detected & Position Correct\\
\midrule
50  events & $82  \pm 9$  & 10/10 & 10/10\\
100 events & $178 \pm 14$ & 10/10 & 10/10\\
200 events & $351 \pm 27$ & 10/10 & 10/10\\
\bottomrule
\end{tabular}
\label{tab:audit}
\end{table}

\subsection{Integration Test Suite}

Table~\ref{tab:tests} summarises the 108-test backend integration suite.
Tests run against an in-memory SQLite backend with Alembic-migrated schema
and monkeypatched Ollama responses; the full suite completes in 43.5\,s.

\begin{table}[t]
\caption{Integration Test Suite Coverage (108 Tests Total)}
\centering\small
\begin{tabular}{lr}
\toprule
Module & Tests\\
\midrule
Authentication \& JWT sessions         & 5\\
Engagement CRUD \& scope validation    & 12\\
Approval workflows                     & 8\\
Tool invocations \& scope checks       & 10\\
Tool executions \& streaming           & 15\\
Findings \& parser suggestions         & 8\\
Report generation                      & 6\\
Audit chain integrity                  & 10\\
Knowledge base (RAG ingestion/search)  & 14\\
Agent pipeline (all phases)            & 20\\
\midrule
\textbf{Total (43.5\,s wall time)}     & \textbf{108}\\
\bottomrule
\end{tabular}
\label{tab:tests}
\end{table}

\subsection{LLM Output Quality and Token Throughput}

Fig.~\ref{fig:tokens} plots mean token counts and throughput
across all planning task types, contrasting the 3B and 14B model roles.
The 3B model handles short-context structured tasks within 80 tokens;
the 14B model produces planning outputs of 320--520 tokens at 12\,tok/s.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  width=\columnwidth, height=4.5cm,
  bar width=8pt,
  ylabel={Output tokens},
  symbolic x coords={Classify, Parse, Recon-Plan, Enum-Plan,
                     Vuln-Map, Exploit-Plan, Write-Finding},
  xtick=data,
  xticklabel style={font=\tiny,rotate=32,anchor=east},
  yticklabel style={font=\scriptsize},
  ylabel style={font=\scriptsize},
  ymin=0, ymax=580,
  nodes near coords,
  nodes near coords style={font=\tiny},
  every node near coord/.append style=
    {/pgf/number format/.cd,fixed,precision=0},
  grid=major, grid style={dotted,gray!40},
  enlarge x limits=0.06,
  legend style={at={(0.98,0.97)},anchor=north east,font=\scriptsize},
]
\addplot+[fill=teal!40,draw=teal!70] coordinates {
  (Classify,62)   (Parse,78)     (Recon-Plan,0)
  (Enum-Plan,0)   (Vuln-Map,0)   (Exploit-Plan,0) (Write-Finding,0)
};
\addplot+[fill=blue!40,draw=blue!70] coordinates {
  (Classify,0)     (Parse,0)       (Recon-Plan,348)
  (Enum-Plan,312)  (Vuln-Map,421)  (Exploit-Plan,518) (Write-Finding,284)
};
\legend{llama3.2:3b, qwen2.5:14b}
\end{axis}
\end{tikzpicture}
\caption{Mean output token counts per task type across both models
  (30 trials each). Zero bars indicate that task type is not assigned
  to that model.}
\label{fig:tokens}
\end{figure}

Finding quality was evaluated by comparing PentAI Pro's generated
findings against a manually authored ground-truth set for three
distinct vulnerability profiles.
Precision was 0.87 (13\% false positives, predominantly
version-based CVE suggestions for patched software) and recall was
0.91 (9\% missed findings, all low-severity).
F$_1 = 0.89$ across the combined finding set.

\subsection{End-to-End Assessment}

Two structured engagements ran against a deliberately vulnerable target
(Apache 2.4.49, nginx 1.18.0, OpenSSH 8.9, outdated TLS configuration).
Five findings were seeded; PentAI Pro identified all five in both runs.
Total pipeline time was 213\,s (run~1) and 228\,s (run~2), with the
difference attributable to nuclei template-load variation.

Real audit chain output from Run~1 (first three events, truncated):

\begin{lstlisting}[style=terminal,caption={Audit chain — first three entries}]
{
  "seq": 1, "event_type": "tool_validation_passed",
  "actor": "agent", "payload": {"tool":"nmap","op":"service_scan"},
  "prev_hash": "0000...0000",
  "evidence_hash": "3a7f9c2e4b1d8a6..."
}
{
  "seq": 2, "event_type": "tool_execution_started",
  "actor": "agent", "payload": {"execution_id": "e8f3..."},
  "prev_hash": "3a7f9c2e4b1d8a6...",
  "evidence_hash": "b12d4f7a903e1c5..."
}
{
  "seq": 3, "event_type": "tool_execution_completed",
  "actor": "system", "payload": {"exit_code": 0},
  "prev_hash": "b12d4f7a903e1c5...",
  "evidence_hash": "c56a2918d04f3b7..."
}
\end{lstlisting}

\subsection{Tool-Vulnerability Coverage Analysis}

The six tools in the PentAI Pro registry address distinct vulnerability
classes. Fig.~\ref{fig:heatmap} quantifies this specialisation as a
coverage matrix, scoring each tool's detection capability for eight
vulnerability classes on a scale of 0 (no coverage) to 3 (primary
detection method). nmap and nuclei provide the broadest cross-class
coverage; dnsx, sslscan, and gobuster are specialist tools with narrow
but authoritative detection of their respective classes.
The agent pipeline's phase-aware tool selection corresponds to
traversing the high-value cells in canonical attack order — beginning
with DNS and port discovery and progressing through service
fingerprinting, CVE exposure, and TLS misconfiguration.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  colorbar,
  colormap/Blues,
  colorbar style={
    width=5pt,
    ytick={0,1,2,3},
    yticklabels={None,Partial,Moderate,Primary},
    yticklabel style={font=\tiny},
  },
  width=\columnwidth, height=5.4cm,
  xtick={0,1,2,3,4,5},
  xticklabels={nmap,nuclei,gobuster,httpx,sslscan,dnsx},
  xticklabel style={font=\tiny,rotate=30,anchor=east},
  ytick={0,1,2,3,4,5,6,7},
  yticklabels={DNS-Recon,Open-Ports,Svc-FP,CVE-Exp,Misconfig,Web-Dir,HTTP-Hdr,TLS},
  yticklabel style={font=\tiny},
  point meta min=0,point meta max=3,
  enlarge x limits=false,
  enlarge y limits=false,
]
\addplot[matrix plot*,mesh/cols=6,point meta=explicit] coordinates {
  (0,0)[1] (1,0)[0] (2,0)[0] (3,0)[0] (4,0)[0] (5,0)[3]
  (0,1)[3] (1,1)[0] (2,1)[0] (3,1)[1] (4,1)[0] (5,1)[0]
  (0,2)[3] (1,2)[1] (2,2)[0] (3,2)[1] (4,2)[1] (5,2)[0]
  (0,3)[1] (1,3)[3] (2,3)[0] (3,3)[0] (4,3)[1] (5,3)[0]
  (0,4)[1] (1,4)[3] (2,4)[1] (3,4)[1] (4,4)[1] (5,4)[0]
  (0,5)[0] (1,5)[2] (2,5)[3] (3,5)[1] (4,5)[0] (5,5)[0]
  (0,6)[1] (1,6)[2] (2,6)[1] (3,6)[3] (4,6)[1] (5,6)[0]
  (0,7)[1] (1,7)[2] (2,7)[0] (3,7)[1] (4,7)[3] (5,7)[0]
};
\end{axis}
\end{tikzpicture}
\caption{Tool--vulnerability coverage matrix (0 = no coverage,
  3 = primary detection). Each tool's specialist domain is visible
  along the high-intensity diagonal; nuclei and nmap provide
  the broadest cross-class coverage.}
\label{fig:heatmap}
\end{figure}

%─────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{Comparison with Related Systems}

Table~\ref{tab:compare} places PentAI Pro in context with the closest
related systems across the properties evaluated in this paper.

\begin{table}[t]
\caption{Feature Comparison with Related Systems}
\centering\scriptsize
\begin{tabular}{lcccc}
\toprule
Feature & PentAI & PentestGPT & OpenVAS & Nuclei\\
         & \textbf{Pro} & \cite{deng2023} & & \cite{nuclei}\\
\midrule
Multi-phase reasoning  & \checkmark & \checkmark & -- & --\\
Local/offline LLM      & \checkmark & --          & N/A & N/A\\
RAG knowledge base     & \checkmark & --          & -- & --\\
Scope enforcement      & 3-layer    & None        & Manual & Manual\\
Tamper-evident audit   & SHA-256    & None        & None & None\\
mTLS inter-node        & \checkmark & --          & -- & --\\
Operator approval flow & \checkmark & Per-step    & -- & --\\
Open-source deployment & \checkmark & --          & \checkmark & \checkmark\\
\bottomrule
\end{tabular}
\label{tab:compare}
\end{table}

\subsection{Scope Safety Guarantees}

The three-layer design provides defence-in-depth rather than a single
authoritative gate. In practice, all out-of-scope attempts were caught
at Layer~1 or Layer~2 before reaching the gateway. Layer~3 exists to
maintain the guarantee even if the command plane is compromised —
an important property for deployments where the Command Node is exposed
to untrusted operators.

\subsection{LLM Quality vs.\ Latency}

CPU-bound inference at 12\,tok/s is the primary performance constraint.
Planning phases that produce 300--500 tokens of structured JSON take
25--40\,s. This is acceptable for an asynchronous background agent
but would be noticeable in interactive use. A discrete GPU with 8\,GB
VRAM would reduce inference time by a factor of 5--10, bringing planning
latency below 5\,s. The architecture supports GPU placement without
code changes.

The quantisation penalty for qwen2.5:14b at Q4\_K\_M versus FP16 was
not measured in this work; published benchmarks from the model authors
report a $<$2\,\% MMLU score drop at this quantisation level~\cite{qwen2024},
a negligible quality cost for the inference-speed gain on consumer hardware.

\subsection{Memory and Disk Footprint}

At steady state the Command Node processes consume 380\,MB for the FastAPI
backend, 512\,MB for the Next.js Node runtime, 480\,MB for the
PostgreSQL 16 instance (with a 200\,MB shared buffer pool), and
9.2\,GB for the loaded qwen2.5:14b model weights.
Total resident memory across all PentAI Pro processes sits at
$\approx$12\,GB, leaving $\approx$13\,GB for the Supabase service
stack (13 Docker containers, $\approx$2.8\,GB combined) and kernel
page cache on the 25\,GB Command Node.
Storage consumption after a 10-engagement run:
the PostgreSQL data directory grew to 1.4\,GB (findings, audit events,
knowledge chunks, execution artifacts), confirming that storage growth
is driven by scan artifacts rather than model weights.

\subsection{Limitations}

The current implementation assesses one host per engagement and does
not model lateral movement across multiple discovered hosts.
The knowledge base requires manual curation; automated ingestion from
NVD or MITRE ATT\&CK feeds is future work.
Prompt injection via tool output remains a theoretical risk despite the
trust-tagging mitigation; a complete solution would require output-format
validation at the model boundary.

%─────────────────────────────────────────────────────────────
\section{Conclusion}

PentAI Pro demonstrates that locally deployed, open-weight language models
can automate the core reasoning steps of a penetration test while
respecting legal scope constraints and producing a forensically
verifiable record. Three findings stand out from this work.

First, three-layer scope enforcement with independent checks at the
LLM planner, backend API, and tool gateway provides robust guarantees
— 150 adversarial injection attempts produced zero out-of-scope executions.
Second, SHA-256 hash chaining is computationally cheap (sub-400\,ms for
200 events) and identifies tampering to single-event granularity.
Third, a 14-billion-parameter quantised model running on a 6-core server
CPU produces planning output of sufficient quality to identify all
pre-seeded findings in structured engagements, suggesting that capable
offensive reasoning no longer requires proprietary cloud inference.

Future directions include GPU-accelerated inference, multi-host lateral
movement planning, automated CVE-feed ingestion, and formal verification
of the scope enforcement invariants.

%─────────────────────────────────────────────────────────────
\balance
\begin{thebibliography}{00}

\bibitem{happe2023}
A.~Happe and J.~Cito,
``Getting pwn'd by AI: Penetration testing with large language models,''
in \textit{Proc.\ ACM Joint European Software Engineering Conf.\ and
  Symp.\ on the Foundations of Software Engineering (ESEC/FSE)},
San Francisco, CA, USA, Dec.\ 2023, pp.\ 2082--2086.

\bibitem{deng2023}
G.~Deng \textit{et al.},
``PentestGPT: An LLM empowered automatic penetration testing tool,''
in \textit{Proc.\ USENIX Security Symp.}, Philadelphia, PA, USA,
Aug.\ 2024, pp.\ 3365--3382.

\bibitem{yao2022}
S.~Yao \textit{et al.},
``ReAct: Synergizing reasoning and acting in language models,''
in \textit{Proc.\ Int.\ Conf.\ on Learning Representations (ICLR)},
Kigali, Rwanda, May 2023.

\bibitem{lewis2020}
P.~Lewis \textit{et al.},
``Retrieval-augmented generation for knowledge-intensive NLP tasks,''
in \textit{Proc.\ Advances in Neural Information Processing Systems
  (NeurIPS)}, vol.\ 33, Dec.\ 2020, pp.\ 9459--9474.

\bibitem{nuclei}
ProjectDiscovery,
``Nuclei: Fast and customisable vulnerability scanner based on simple
  YAML based DSL,''
\url{https://github.com/projectdiscovery/nuclei}, 2024.

\bibitem{nessus}
Tenable, Inc.,
``Nessus professional,''
\url{https://www.tenable.com/products/nessus}, 2024.

\bibitem{qwen2024}
Qwen Team,
``Qwen2.5 technical report,''
\textit{arXiv preprint arXiv:2412.15115}, Dec.\ 2024.

\bibitem{andriesse2018}
D.~Andriesse \textit{et al.},
``An in-depth analysis of disassembly on full-scale x86/x64 binaries,''
in \textit{Proc.\ USENIX Security Symp.}, Austin, TX, USA,
Aug.\ 2016, pp.\ 583--600.

\bibitem{supabase}
Supabase Inc.,
``Supabase: The open source Firebase alternative,''
\url{https://supabase.com}, 2024.

\bibitem{owasp}
OWASP Foundation,
``OWASP Testing Guide v4.2,''
\url{https://owasp.org/www-project-web-security-testing-guide/}, 2021.

\end{thebibliography}

\end{document}
